---
title: | 
  <div style="text-align: center; color: black; font-size: 35px; font-weight: bold;">[EM1413] - Data loading and cleaning</div>
subtitle: | 
  <div style="text-align: center; color: black; font-size: 30px;">BRFSS 2024 Analysis</div>
author: |
  <div style="text-align: center; color: black; font-size: 22px; font-weight: bold;">Group's name</div>
  <div style="text-align: left; margin: 0 auto; width: fit-content; color: black; font-size: 20px;">
    • Riccardo Gasparini – 874762<br>
    • Gianluca Michelon – <br>
    • Elisa Pascon – 891393<br>
    • Samuele Pasin – 
  </div>
date: |
  <div style="text-align: center; color: black; font-size: 17px;">a.y. 2025/2026</div>
  <div style="text-align: center; color: black; font-size: 16px; font-style: italic">exam call: 25th May 2026</div>
output:
  html_document:
    theme: sandstone
    toc: TRUE
    toc_float:
      smooth_scroll: TRUE
    toc_depth: 2
    number_sections: TRUE
    fig_caption: yes
    code_folding: show
    highlight: pygments
    self_contained: TRUE
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
pacman::p_load(foreign, tidyverse, forcats, survey, srvyr, lme4, lmerTest, knitr, 
               kableExtra, gt, gtsummary, scales, patchwork, ggeffects, palettes, ggcorrplot, GGally)
```

# Research question

> What demographic, socioeconomic and behavioral factors predict participation in preventive cancer screening?

### Source

Dataset: [https://www.cdc.gov/brfss/annual_data/annual_2024.html](https://www.cdc.gov/brfss/annual_data/annual_2024.html)

# Loading and preparing the data

```{r sourcing-loading-file}
brfss24_sub <- read.csv("brfss24_sub.csv") # from BRFSS_ne_data_prep.R

brfss24_sub <- brfss24_sub %>%
  mutate(across(
    .cols = c(
      where(is.integer),
      where(is.character),
      where(is.logical),
      -X_PSU,
      -X_STSTR,
      -X_LLCPWT
    ),
    .fns = as.factor
  ))
```

Creating a survey object in order to properly apply weights:

```{r svy-obj}
brfss.design <- brfss24_sub %>%
  as_survey_design(
    ids = X_PSU,
    weights = X_LLCPWT,
    strata = X_STSTR,
    nest = TRUE
  )
options(survey.lonely.psu = "adjust")
```

## Taking a glimpse of what the data looks like

```{r glimpse}
glimpse(brfss.design$variables)
```

### Summary

```{r summary}
brfss.design %>%
  tbl_svysummary(
    include  = c(X_AGE_G, X_SEX, X_EDUCAG, X_INCOMG1, X_HLTHPL2),
    label    = list(
      X_AGE_G   ~ "Age group",
      X_SEX     ~ "Sex",
      X_EDUCAG  ~ "Education",
      X_INCOMG1 ~ "Household income",
      X_HLTHPL2 ~ "Health insurance"
    ),
    statistic = list(all_categorical() ~ "{p}% ({n_unweighted})")
  ) %>%
  modify_caption("**Weighted demographic characteristics, BRFSS 2024**")
```

# Exploratory data analysis

```{r intro-corr}
corr_data <- brfss.design$variables %>%
  select(any_screening, EXERANY2, X_RFSMOK3, X_BMI5, X_EDUCAG, EMPLOY1, RENTHOM1, X_RACE, X_INCOMG1, X_HLTHPL2, X_SEX) %>%
  rename(
    "Any Screening" = any_screening,
    "Exercise"      = EXERANY2,
    "Smoking"       = X_RFSMOK3,
    "BMI"           = X_BMI5,
    "Education"     = X_EDUCAG,
    "Employment"    = EMPLOY1,
    "Renting Home"  = RENTHOM1,
    "Race"          = X_RACE,
    "Income"        = X_INCOMG1,
    "Has Insurance" = X_HLTHPL2,
    "Sex"           = X_SEX
  ) %>%
  mutate(across(everything(), ~ as.numeric(as.factor(.)))) %>%
  na.omit()

corr_matrix <- cor(corr_data)

ggcorrplot(
  corr_matrix,
  method = "square",
  type = "lower",
  lab = TRUE,
  lab_size = 3,
  colors = c("tomato2", "white", "steelblue"),
  title = "Correlation Matrix of Predictors",
  ggtheme = theme_minimal()
) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

## Demographic features

### Age patterns

```{r age-patterns}
agesex_screen <- brfss.design %>%
  filter(!is.na(HADSIGM4), !is.na(X_AGE_G), !is.na(X_SEX)) %>%
  group_by(age = X_AGE_G,
           sex = X_SEX,
           response = HADSIGM4) %>%
  summarise(prop = survey_mean(vartype = "ci"), .groups = "drop") %>%
  filter(response == 1)

ggplot(agesex_screen, aes(age, prop, colour = sex, group = sex)) +
  geom_line(linewidth = 1) +
  # geom_ribbon(aes(ymin = prop_low, ymax = prop_upp, fill = sex),
  #             alpha = 0.15,
  #             colour = NA) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format()) +
  scale_colour_manual(values = c("forestgreen", "darkmagenta")) +
  scale_fill_manual(values   = c("forestgreen", "darkmagenta")) +
  labs(
    title  = "Colorectal cancer screening by age and sex (weighted)",
    x      = "Age group",
    y = "Proportion screened",
    colour = "Sex",
    fill = "Sex"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1),
        plot.title = element_text(face = "bold", size = 14))
```

```{r age-distribution}
ggplot(brfss24_sub %>% drop_na(X_AGE_G)) +
  geom_bar(
    aes(X_AGE_G, fill = X_AGE_G),
    col = "black",
    linewidth = 0.2,
    alpha = .8
  ) +
  labs(x = "Age", y = "Count", title = "Age distribution") +
  # scale_fill_brewer(palette = "YlOrRd") +
  scale_color_viridis_d() +
  theme_bw() +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 14)) 
```

### Sex

```{r sex-distribution}
plot_don <- brfss24_sub %>%
  group_by(X_SEX) %>%
  summarise(count = n()) %>%
  mutate(
    prop = count / sum(count),
    label = percent(prop)
  )

ggplot(plot_don, aes(x = "", y = prop, fill = X_SEX)) +
  geom_bar(stat = "identity", width = 1, alpha = .8, color = "black", linewidth =.2) +
  coord_polar("y", start = 0) +
  # 3. Add the labels
  geom_text(aes(label = label), 
            position = position_stack(vjust = 0.5), 
            size = 5, 
            color = "white",
            fontface = "bold") +
  scale_fill_manual(values = c("Male" = "darkblue", "Female" = "red3")) +
  theme_void() +
  labs(title = "Sex Distribution Percentage", fill = "Sex") +
  theme(plot.title = element_text(face = "bold", size = 14))
```

### BMI distribution

```{r bmi-distribution}
avg_BMI <- brfss24_sub %>% drop_na(X_BMI5) %>% summarise(mean_BMI = mean(X_BMI5))

brfss24_sub %>% drop_na(X_BMI5) %>%
  ggplot() +
  geom_histogram(aes(x = X_BMI5), binwidth = 1.5, fill = "darkorange", colour = 'black', alpha = .8, linewidth = .2) +
  labs(title = "Distribution of BMI", x = 'BMI', y = 'count') +
  theme_minimal() +
  geom_vline(aes(xintercept = avg_BMI[[1]]), colour = 'red')
```


### BMI relationship

```{r bmi-screening}
bmi_plot_data <- brfss.design$variables %>%
  select(any_screening, X_BMI5) %>%
  filter(!is.na(any_screening), !is.na(X_BMI5)) %>%
  mutate(screening_label = case_when(any_screening == "1" ~ "Screened", TRUE ~ "Not Screened"))

ggplot(bmi_plot_data,
       aes(x = screening_label, y = X_BMI5, fill = screening_label)) +
  geom_violin(alpha = 0.6, color = "black") +
  geom_boxplot(
    width = 0.15,
    fill = "white",
    color = "black",
    outlier.alpha = 0.3
  ) +
  scale_fill_manual(values = c("brown", "yellow")) +
  labs(title = "BMI Distribution by Cancer Screening Status", x = "Participation in Any Preventive Screening", y = "Body Mass Index (BMI)") +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11)
  )
```


## Socioeconomic features

### Income distribution

```{r barplot-income, fig.align='center'}
ggplot(brfss24_sub %>% drop_na(X_INCOMG1)) +
  geom_bar(
    aes(X_INCOMG1, fill = X_INCOMG1),
    col = "black",
    linewidth = 0.2,
    alpha = .8
  ) +
  labs(x = "Income", y = "Count", title = "Income distribution") +
  scale_fill_brewer(palette = "YlOrRd") +
  theme_bw() +
  theme(legend.position = "none", 
        plot.title = element_text(face = "bold", size = 14)) 
```

```{r barplot-income-relative}
brfss.design %>%
  filter(!is.na(X_INCOMG1)) %>%
  group_by(X_INCOMG1) %>%
  summarise(prop = survey_mean(vartype = "ci")) %>%
  ggplot(aes(X_INCOMG1, prop, fill = X_INCOMG1)) +
  geom_col(col = "black",
           linewidth = 0.2,
           alpha = 0.8) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Income", y = "Weighted proportion", title = "Income distribution (weighted)") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14))
```

### Employment distribution

```{r barplot-employment, fig.align='center'}
ggplot(brfss24_sub %>% drop_na(EMPLOY1)) +
  geom_bar(
    aes(EMPLOY1, fill = EMPLOY1),
    col = "black",
    linewidth = 0.2,
    alpha = .8
  ) +
  labs(x = "Income", y = "Count", title = "Employment distribution") +
  scale_fill_brewer(palette = "YlOrRd") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(
          angle = 45,
          hjust = 1,
          size = 10
        ),
        plot.title = element_text(face = "bold", size = 14))
```

### Education distribution

```{r educ-distribution}
ggplot(brfss24_sub %>% drop_na(X_EDUCAG),
       aes(X_EDUCAG, fill = X_EDUCAG)) +
  geom_bar(col = "black",
           linewidth = 0.2,
           alpha = .8) +
  scale_x_discrete(labels = label_wrap(25)) +
  labs(x = "Education level", y = "Count", title = "Education level distribution") +
  scale_fill_brewer(palette = "YlOrRd") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(
          angle = 45,
          hjust = 1,
          size = 10
        ),
        plot.title = element_text(face = "bold", size = 14))
```

## Behavioural factors

```{r smoke-ex-screen}
heatmap_data <- brfss.design %>%
  mutate(
    EXERANY2 = case_when(
      EXERANY2 == "1" ~ "Yes",
      EXERANY2 %in% c("0", "2") ~ "No",
      TRUE ~ NA_character_
    ),
    X_RFSMOK3 = case_when(
      X_RFSMOK3 == "1" ~ "No",
      X_RFSMOK3 %in% c("2") ~ "Yes",
      TRUE ~ NA_character_
    )
  ) %>%
  group_by(exercise = EXERANY2,
           smoking = X_RFSMOK3,
           screened = any_screening) %>%
  summarise(prop = survey_mean(vartype = "ci"), .groups = "drop") %>%
  filter(screened == "1", !is.na(exercise), !is.na(smoking))

ggplot(heatmap_data, aes(x = exercise, y = smoking, fill = prop)) +
  geom_tile(linewidth = 0.2, alpha = .9, color = "black") +
  geom_text(
    aes(label = percent(prop, accuracy = 0.1)),
    color = "white",
    fontface = "bold",
    size = 5
  ) +
  scale_fill_gradient(
    low = "lightblue",
    high = "darkblue",
    labels = percent_format()
  ) +
  labs(
    title = "Cancer Screening Rates by Smoking & Exercise Habits",
    subtitle = "Survey-weighted proportions",
    x = "Participated in Any Exercise (Past Month)",
    y = "Smoking Status",
    fill = "Screened (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    axis.title = element_text(face = "bold")
  )
```

```{r income-screening}
income_plot_data <- brfss.design %>%
  mutate(
    income = as.character(X_INCOMG1),
    screening = case_when(
      any_screening == "1" ~ "Screened",
      any_screening %in% c("0", "2") ~ "Not Screened",
      TRUE ~ NA_character_
    )
  ) %>%
  group_by(income, screening) %>%
  summarise(prop = survey_mean(vartype = "ci"), .groups = "drop") %>%
  filter(!is.na(income), !is.na(screening))

ggplot(income_plot_data, aes(x = income, y = prop, fill = screening)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7,
    color = "black",
    linewidth = 0.2, 
    alpha = .9
  ) +
  scale_fill_manual(values = c(
    "Screened" = "navyblue",
    "Not Screened" = "green4"
  )) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Preventive Cancer Screening Rates by Income Bracket",
    subtitle = "Survey-weighted proportions with 95% confidence intervals",
    x = "Annual Household Income Bracket",
    y = "Proportion of Population",
    fill = "Screening Status"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      size = 10
    )
  )
```


## Health insurance coverage

```{r insurance-screening-summary}
calc_sigm <- brfss.design %>%
  group_by(insurance = X_HLTHPL2, response = HADSIGM4) %>%
  summarise(prop = survey_mean(na.rm = TRUE, vartype = "ci"),
            .groups = "drop") %>%
  filter(response == "1", !is.na(insurance), !is.na(response)) %>%
  mutate(screening = "HADSIGM4")

calc_mam <- brfss.design %>%
  group_by(insurance = X_HLTHPL2, response = HADMAM) %>%
  summarise(prop = survey_mean(na.rm = TRUE, vartype = "ci"),
            .groups = "drop") %>%
  filter(response == "1", !is.na(insurance), !is.na(response)) %>%
  mutate(screening = "HADMAM")

calc_psa <- brfss.design %>%
  group_by(insurance = X_HLTHPL2, response = PSATEST1) %>%
  summarise(prop = survey_mean(na.rm = TRUE, vartype = "ci"),
            .groups = "drop") %>%
  filter(response == "1", !is.na(insurance), !is.na(response)) %>%
  mutate(screening = "PSATEST1")

ins_screen <- bind_rows(calc_sigm, calc_mam, calc_psa)

ins_screen %>%
  select(insurance, screening, prop) %>%
  pivot_wider(names_from = screening, values_from = prop)
```

```{r insurance-screening-graph}
ins_screen %>%
  mutate(
    screening = recode(
      screening,
      HADSIGM4 = "Colonoscopy/sigmoid.",
      HADMAM   = "Mammogram",
      PSATEST1 = "PSA test"
    )
  ) %>%
  ggplot(aes(insurance, prop, fill = insurance)) +
  geom_col(
    position = "dodge",
    alpha = 0.8,
    colour = "black",
    linewidth = 0.2
  ) +
  facet_wrap( ~ screening) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("darkviolet", "skyblue")) +
  labs(title = "Screening participation by health insurance status",
       x = NULL,
       y = "Weighted proportion (%)",
       fill = NULL) +
  theme_bw() +
  theme(legend.position = "none")
```

# Models

```{r model-var-selection}
# da aggiungere prima
# model_data <- brfss24_sub %>%
#   filter(!is.na(any_screening)) %>%
#   drop_na(X_AGE_G, X_SEX, X_RACE, MARITAL, VETERAN3, X_INCOMG1, X_EDUCAG, EMPLOY1, RENTHOM1, X_HLTHPL2,
#           PERSDOC3, MEDCOST1, X_RFSMOK3, EXERANY2, X_BMI5, X_RFDRHV9)
# 
# brfss.design.clean <- model_data %>%
#   as_survey_design(ids = X_PSU, strata = X_STSTR, weights = X_LLCPWT, nest = TRUE)
# 
# svy_null <- svyglm(any_screening ~ 1, design = final_design, family = quasibinomial())
# 
# svy_full <- svyglm(
#   any_screening ~ X_AGE_G + X_SEX + X_RACE + MARITAL + VETERAN3 + X_INCOMG1 + X_EDUCAG + EMPLOY1 +
#     RENTHOM1 +  X_HLTHPL2 + PERSDOC3 + MEDCOST1 + X_RFSMOK3 + EXERANY2 + X_BMI5 + X_RFDRHV9, 
#   design = final_design, 
#   family = quasibinomial(link = "logit")
# )
# 
# svy_forward <- step(
#   svy_null,
#   scope = list(lower = svy_null, upper = svy_full),
#   direction = "forward",
#   trace = 1
# )
# 
# svy_backward <- step(
#   svy_full,
#   direction = "backward",
#   trace = 1
# )
# 
# svy_stepwise <- step(
#   svy_null,
#   scope = list(lower = svy_null, upper = svy_full),
#   direction = "both",
#   trace = 1
# )
# 
# formula(svy_forward)
# formula(svy_backward)
# formula(svy_stepwise)
# ----- NON FUNZIONA -----
```

## Modello random intercept

```{r}

library(lme4)

model_mixed_female <- glmer(
  any_screening ~ X_AGE_G + X_EDUCAG + X_INCOMG1 +
                  X_HLTHPL2 + PERSDOC3 + X_RFSMOK3 +
                  EXERANY2 + X_BMI5 + X_RACE +
                  (1 | X_STATE),
  family  = binomial,
  data    = brfss24_sub %>% filter(X_SEX == "Female"),
  control = glmerControl(
    optimizer = "bobyqa",
    optCtrl   = list(maxfun = 2e5)
  )
)

summary(model_mixed_female)

sjPlot::plot_model(model_mixed_female, type = "re", sort.est = TRUE, show.values = TRUE,value.offset = 0.3)

sjPlot::plot_model(model_mixed_female, show.values = TRUE, show.p=TRUE, value.offset = 0.5, value.size = 3.5)

# Calcolo dell'ICC
# L'ICC indica quale frazione della variabilità totale non spiegata dai predittori è attribuibile alle differenze tra stati.
# L'ICC risulta molto basso. Solo lo 0,7 della varaibilità dipende dallo stato

sigma2_u <- as.data.frame(VarCorr(model_mixed_female))$vcov[1]
icc <- sigma2_u / (sigma2_u + pi^2/3)
icc

```



## Prova modello 1 - Logistic regression

```{r model1-data}
model_svyglm <- svyglm(
  any_screening ~ X_AGE_G + X_SEX + X_EDUCAG + X_INCOMG1 +
    X_HLTHPL2 + X_RFSMOK3 + EXERANY2 + X_BMI5 + X_RACE,
  design = brfss.design,
  family = quasibinomial(link = "logit")
)

summary(model_svyglm)
```

- Sex (`X_SEXFemale`): strongest predictor ($Estimate = 2.93$). Being female massively increases the odds of having had "any screening" compared to males $\rightarrow$ however, this is due to the fact that there are two screening that are exclusively for women;
- Age (`X_AGE_G.L`): (.L = "Linear" trend). The positive estimate ($4.17$) means that as age increases, the likelihood of screening increases significantly;
- Insurance (`X_HLTHPL2Not insured`): negative estimate ($-0.507$) shows that being uninsured is a significant barrier, decreasing the odds of screening compared to those with insurance;
- BMI (`X_BMI5`): for every unit increase in BMI, the odds of screening increase slightly ($0.025$);the number is small, the p-value is highly significant ($***$) $\rightarrow$ from the EDA grphs, though, we can see there is no much correlation between the two variables;


- Asian only: significantly less likely to have had screening ($-0.807$);
- Hispanic: significantly less likely to have had screening ($-0.202$);
- Black / Multiracial: there is no significant difference compared to the reference group ($p > 0.05$).


- Education (`X_EDUCAG.L`): there is a significant positive linear trend. Higher education levels correlate with a higher likelihood of screening;
- Income: most income brackets aren't significantly different from the reference (lowest income), except for the \$100k–\$200k group, which shows a significant increase in screening odds.


```{r m1-results}
(
  tbl_svyglm <- tbl_regression(
    model_svyglm,
    exponentiate = TRUE,
    # gives odds ratios
    label = list(
      # just replaces labels
      X_AGE_G ~ "Age group",
      X_SEX ~ "Sex",
      X_EDUCAG ~ "Education",
      X_INCOMG1 ~ "Household income",
      X_HLTHPL2 ~ "Health insurance",
      X_RFSMOK3 ~ "Current smoker",
      EXERANY2 ~ "Any exercise",
      X_BMI5 ~ "BMI",
      X_RACE ~ "Race/Ethnicity"
    )
  ) %>%
    bold_p(t = 0.05) %>% # significant = bold
    bold_labels() %>%
    modify_caption("**Survey-weighted logistic regression: OR (95% CI)**")
)
# if the CI contains 1 = not significant
```


```{r m1-predict-insurance}
ggeffects::ggpredict(model_svyglm, terms = c("X_INCOMG1", "X_HLTHPL2")) %>%
  plot() +
  labs(
    title = "Predicted probability of screening by income & insurance",
    x = "Income bracket",
    y = "Predicted probability",
    color = "Has insurance"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 13)
  )
```

```{r}
ggeffects::ggpredict(model_svyglm, terms = c("X_INCOMG1", "X_EDUCAG")) %>%
  plot() +
  labs(
    title = "Predicted probability of screening by income & insurance",
    x = "Income bracket",
    y = "Predicted probability",
    color = "Has insurance"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "bottom"
  )
```


Looking at the dots: across every single income bracket, people with insurance have a higher predicted probability of being screened than those without.

# References

- [https://ikwak2.github.io/tutorials/BRFSS/BRFSS.html](https://ikwak2.github.io/tutorials/BRFSS/BRFSS.html)

## Useful

To take a look:

- Lumley T (2023). *Complex Surveys: A Guide to Analysis Using R*. Wiley;
- Lumley T. `survey` package documentation: <http://r-survey.r-forge.r-project.org/survey/>;
- Bates D et al. `lme4`: Linear mixed-effects models using Eigen and S4. *J Stat Softw* 67(1), 2015;
- Tutorial on BRFSS in R: <https://ikwak2.github.io/tutorials/BRFSS/BRFSS.html>.
