---
title: "Data loading and cleaning - BRFSS"
output: 
  html_document: 
    theme: flatly #sandstone #spacelab #flatly
    highlight: pygments
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research question

> What demographic, socioeconomic and behavioral factors predict participation in preventive cancer screening?

### Source

Dataset: [https://www.cdc.gov/brfss/annual_data/annual_2024.html](https://www.cdc.gov/brfss/annual_data/annual_2024.html)

## Package loading

```{r packages, include=FALSE}
pacman::p_load(RCurl,
               forcats,
               foreign,
               downloader,
               survey,
               srvyr,
               tidyverse,
               scales,
               palettes,
               knitr,
               kableExtra)
```

## Data loading

```{r data-loading}
brfss_complete <- read.xport('BRFSS_2024.XPT')

sprintf("BRFSS data have %d records and %d number of variables",
        nrow(brfss_complete),
        ncol(brfss_complete))

brfss_complete[1:5, 1:10] # just to get familiar with the dataset
```
### Interesting variables

#### Demographic features

- `_STATE`: US State
- `_AGE_G`: Six-level imputed age category
- `_SEX`: Sex variable
- `_RACE`: Race/ethnicity categories
- `MARITAL`: marital status
- `VETERAN3`: Have you ever served on active duty in the United States Armed Forces, either in the regular military or in a National Guard or military reserve unit?

#### Cancer screening participation

Absolute answers:

- `CERVSCRN`: Have you ever had a cervical cancer screening test?
    - `CRVCLPAP`: At your most recent cervical cancer screening, did you have a Pap test?
    - `CRVCLHPV`: At your most recent cervical cancer screening, did you have an H.P.V. test?
- `HADMAM`: Have you ever had a mammogram?
- `HADSIGM4`: Colonoscopy and sigmoidoscopy are exams to check for colon cancer. Have you ever had either of these exams?
- `SMALSTOL`: One stool test uses a special kit to obtain a small amount of stool at home and returns the kit to the doctor or the lab. Have you ever had this test?
- `LCSCTSC1`: Have you ever had a CT or CAT scan of your chest area?
- `PSATEST1`: Have you ever had a P.S.A. test?

Including recency:

- `_RFPAP37`: [combined] Female respondents with and intact cervix, aged 21-65, who have had a pap test in the past three years
- `_PAPHPV1`: [combined] Female respondents with an intact cervix, aged 21-65 that have had a pap test in last 3 years or aged 30-65 that have had an HPV test in last 5 years
- `_HPV5YR1`: [combined] Female respondents with and intact cervix, aged 30-65, who have had an HPV test in the past five years
- `_RFMAM23`: Women respondents aged 40+ who have had a mammogram in the past two years
- `_MAM402Y`: Women respondents aged 40-74 who have had a mammogram in the past two years

#### Socioeconomic factors

- `_INCOMG1`: Income categories (better grouping)
- `_EDUCAG`: Level of education completed (better grouping) [alternative `EDUCA`]
- `EMPLOY1`: Employment status
- `RENTHOM1`: Do you own or rent your home?
- [maybe too complicated to include them] `FOODSTMP` / `SDHFOOD1` / `SDHBILLS` / `SDHUTILS` / `SDHTRNSP`: food stamps, food insecurity, unable to pay bills, utility cutoff, transportation barriers

#### Insurance coverage

- `_HLTHPL2`: Adults who had some form of health insurance
  - `_HCVU654`: Respondents aged 18-64 who have any form of health insurance
- `PERSDOC3`: Do you have one person or a group of doctors that you think of as your personal health care provider?
- `MEDCOST1`: Was there a time in the past 12 months when you needed to see a doctor but could not because you could not afford it?

#### Behavioural

- `_RFSMOK3`: Adults who are current smokers
- `EXERANY2`: During the past month, other than your regular job, did you participate in any physical activities or exercises such as running, calisthenics, golf, gardening, or walking for exercise?
- `_BMI5`: BMI (continuous)
- `_RFDRHV9`: Heavy drinkers (adult men having more than 14 drinks per week and adult women having more than 7 drinks per week)

#### Other

- `HADHYST2`: Have you had a hysterectomy? (to exclude from cervical screening)

#### Weights and data

- `_PSU`: Primary Sampling Unit (Equal to Annual Sequence Number)
- `_LLCPWT`: Final weight assigned to each respondent: Land-line and cell-phone data  (Raking derived weight)
- `_STSTR`: Sample Design Stratification Variable

```{r selecting}
# to be refined
sel_variables <- c(
  "X_PSU", "X_STSTR", "X_LLCPWT", # Weights
  "X_STATE", "X_AGE_G", "X_SEX", "X_RACE", "MARITAL", "VETERAN3", # Demographics
  "CERVSCRN", "CRVCLPAP", "CRVCLHPV", "HADMAM", "HADSIGM4", # Cancer screening (absolute)
  "SMALSTOL", "LCSCTSC1", "PSATEST1",
  "X_RFPAP37", "X_PAPHPV1", "X_HPV5YR1", "X_RFMAM23", "X_MAM402Y", # Cancer screening (recency / combined)
  "X_INCOMG1", "X_EDUCAG", "EMPLOY1", "RENTHOM1", # Socioeconomic
  "X_HLTHPL2", "X_HCVU654", "PERSDOC3", "MEDCOST1", # Insurance
  "X_RFSMOK3", "EXERANY2", "X_BMI5", "X_RFDRHV9" # Behavioral
)

brfss24 <- brfss_complete %>%
  select(any_of(sel_variables))
```

### Data cleaning

Cleaning and re-levelling the data:

```{r state-labels, include=FALSE}
state_labels <- c(
  "1" = "Alabama",
  "2" = "Alaska",
  "4" = "Arizona",
  "5" = "Arkansas",
  "6" = "California",
  "8" = "Colorado",
  "9" = "Connecticut",
  "10" = "Delaware",
  "11" = "District of Columbia",
  "12" = "Florida",
  "13" = "Georgia",
  "15" = "Hawaii",
  "16" = "Idaho",
  "17" = "Illinois",
  "18" = "Indiana",
  "19" = "Iowa",
  "20" = "Kansas",
  "21" = "Kentucky",
  "22" = "Louisiana",
  "23" = "Maine",
  "24" = "Maryland",
  "25" = "Massachusetts",
  "26" = "Michigan",
  "27" = "Minnesota",
  "28" = "Mississippi",
  "29" = "Missouri",
  "30" = "Montana",
  "31" = "Nebraska",
  "32" = "Nevada",
  "33" = "New Hampshire",
  "34" = "New Jersey",
  "35" = "New Mexico",
  "36" = "New York",
  "37" = "North Carolina",
  "38" = "North Dakota",
  "39" = "Ohio",
  "40" = "Oklahoma",
  "41" = "Oregon",
  "42" = "Pennsylvania",
  "44" = "Rhode Island",
  "45" = "South Carolina",
  "46" = "South Dakota",
  "47" = "Tennessee",
  "48" = "Texas",
  "49" = "Utah",
  "50" = "Vermont",
  "51" = "Virginia",
  "53" = "Washington",
  "54" = "West Virginia",
  "55" = "Wisconsin",
  "56" = "Wyoming",
  "66" = "Guam",
  "72" = "Puerto Rico",
  "78" = "Virgin Islands"
)
```


```{r clean-relevel}
brfss24 <- brfss24 %>%
  mutate(across(-X_RACE, ~ na_if(., 7))) %>% # transforming 7, 9, 77 and 99 values into NAs
  mutate(across(everything(), ~ na_if(., 9))) %>%
  mutate(across(everything(), ~ na_if(., 77))) %>%
  mutate(across(everything(), ~ na_if(., 99))) %>%
  mutate(
    X_STATE = factor(
      as.character(X_STATE),
      levels = names(state_labels), # hidden chunk in knitted file
      labels = state_labels
    ),
    X_SEX = factor(
      X_SEX,
      levels = c(1, 2),
      labels = c("Male", "Female")
    ),
    MARITAL = factor(
      MARITAL,
      levels = 1:6,
      labels = c(
        "Married",
        "Divorced",
        "Widowed",
        "Separated",
        "Never married",
        "Unmarried couple"
      )
    ),
    VETERAN3 = factor(
      VETERAN3,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    X_AGE_G = factor(
      X_AGE_G,
      levels = 1:6,
      labels = c(
        "Age 18 to 24",
        "Age 25 to 34",
        "Age 35 to 44",
        "Age 45 to 54",
        "Age 55 to 64",
        "Age 65 or older"
      ),
      ordered = TRUE
    ),
    X_INCOMG1 = factor(
      X_INCOMG1,
      levels = 1:7,
      labels = c(
        "<15k",
        "15–25k",
        "25–35k",
        "35–50k",
        "50–100k",
        "100–200k",
        "200k+"
      )
    ),
 
    X_EDUCAG = factor(X_EDUCAG, ordered = TRUE),
    EMPLOY1  = factor(EMPLOY1),
    RENTHOM1 = factor(RENTHOM1),
    X_HLTHPL2 = factor(
      X_HLTHPL2,
      levels = c(1, 2),
      labels = c("Insured", "Not insured")
    ),
    PERSDOC3 = factor(
      PERSDOC3,
      levels = c(1, 2, 3),
      labels = c(
        "Yes, one",
        "Yes, more than one",
        "No"
      )
    ),
    MEDCOST1 = factor(
      MEDCOST1,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    X_HCVU654 = factor(
      X_HCVU654,
      levels = c(1, 2),
      labels = c(
        "Yes",
        "No"
      )
    ),
    X_RFSMOK3 = factor(X_RFSMOK3),
    EXERANY2 = factor(
      EXERANY2,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    X_RFDRHV9 = factor(X_RFDRHV9),
    X_BMI5 = as.numeric(X_BMI5),
    X_RACE = factor(
      X_RACE,
      levels = 1:8,
      labels = c(
        "White only, non-Hispanic",
        "Black only, non-Hispanic",
        "American Indian or Alaskan Native only, non-Hispanic",
        "Asian only, non-Hispanic",
        "Native Hawaiian or other Pacific Islander only, non-Hispanic",
        "Other race only, non-Hispanic",
        "Multiracial, non-Hispanic",
        "Hispanic"
      )
    ),
    CERVSCRN = factor(
      CERVSCRN,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    CRVCLPAP = factor(
      CRVCLPAP,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    CRVCLHPV = factor(
      CRVCLHPV,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    X_RFPAP37 = factor(
      X_RFPAP37,
      levels = c(1, 2),
      labels = c(
        "Had PAP test in last 3 years",
        "Had PAP test more than 3 years ago or have not had one"
      )
    ),
    X_PAPHPV1 = factor(
      X_PAPHPV1,
      levels = c(1, 2),
      labels = c(
        "Had PAP test in last 3 years or HPV test in last 5 years",
        "Have not had PAP test in last 3 years and HPV test in last 5 years"
      )
    ),
    X_HPV5YR1 = factor(
      X_HPV5YR1,
      levels = c(1, 2),
      labels = c(
        "Had HPV test in last 5 years",
        "Had HPV test more than 5 years ago or have not had one"
      )
    ),
    HADMAM = factor(
      HADMAM,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    X_RFMAM23 = factor(
      X_RFMAM23,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    X_MAM402Y = factor(
      X_MAM402Y,
      levels = c(1, 2),
      labels = c(
        "Yes",
        "No"
      )
    ),
    HADSIGM4 = factor(
      HADSIGM4,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    SMALSTOL = factor(
      SMALSTOL,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    LCSCTSC1 = factor(
      LCSCTSC1,
      levels = c(1, 2),
      labels = c("Yes", "No")
    ),
    PSATEST1 = factor(
      PSATEST1,
      levels = c(1, 2),
      labels = c("Yes", "No")
    )
 
  )
```

If we had to convert it into a survey design object:

```{r svy-obj}
brfss.design <- brfss24 %>%
  select(any_of(sel_variables)) %>%
  as_survey_design(
    ids = X_PSU,
    weights = X_LLCPWT,
    strata = X_STSTR,
    nest = TRUE
  )
```

### Taking a glimpse of what the data looks like

```{r glimpse}
glimpse(brfss24)
```

Summary:

```{r summary}
brfss24 %>%
  select(-X_PSU, -X_STSTR, -X_LLCPWT) %>%
  summary()
```

## Exploratory analysis

```{r barplot-income, fig.align='center'}
ggplot(brfss24 %>% drop_na(X_INCOMG1)) +
  geom_bar(
    aes(X_INCOMG1, fill = X_INCOMG1),
    col = "black",
    linewidth = 0.2,
    alpha = .8
  ) +
  labs(x = "Income", y = "Dollars ($)", title = "Levels of income through the whole sample") +
  scale_fill_palette_d(met_palettes$Lakota) +
  theme_bw() +
  theme(legend.position = "none") 
```

```{r screening-proportions}
scree_part <- brfss24 %>%
  select(CERVSCRN,
         CRVCLPAP,
         CRVCLHPV,
         HADMAM,
         HADSIGM4,
         SMALSTOL,
         LCSCTSC1,
         PSATEST1) %>%
  pivot_longer(cols = everything(),
               names_to = "screening",
               values_to = "response") %>%
  filter(!is.na(response)) %>%
  group_by(screening) %>%
  summarise(count = n(), perc_yes = round(mean(response == "Yes"), 4) *
              100)

kable(scree_part, format = "html", caption = "Screening Participation") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, "Summary Statistics" = 2))
```

## References

- [https://ikwak2.github.io/tutorials/BRFSS/BRFSS.html](https://ikwak2.github.io/tutorials/BRFSS/BRFSS.html)
